---
title: "Tutorial: naive Bayes in Pakcage e1071 for Titanic Prediction"
author: "Subbu K"
date: "11/21/2019"
output: word_document
---

This is a tutorial on using the naive Bayes algorithm in the e1071 package to predict Titanic survivors.

# prepare data

First load the training data in csv format, and then convert "Survived" to nominal variable and "Pclass" to ordinal variable.
```{r}
trainset <- read.csv("C:/Users/Administrator/Documents/MSDatascience/IST707/Week7/itanic-train.csv")
trainset$Survived=factor(trainset$Survived)
trainset$Pclass=ordered(trainset$Pclass)
```

Then load the test data and convert attributes in similar way.
```{r}
testset <- read.csv("C:/Users/Administrator/Documents/MSDatascience/IST707/Week7/titanic-test.csv")
testset$Survived=factor(testset$Survived)
testset$Pclass=ordered(testset$Pclass)
```

Then remove some attributes that are not likely to be helpful, such as "embarked" - create a new data set with all other attributes. Process the train and test set in the same way. 
```{r}
myVars=c("Pclass", "Sex", "Age", "SibSp", "Fare", "Survived")
newtrain=trainset[myVars]
newtest=testset[myVars]
```
# naive Bayes in e1071

Now load the package e1071
```{r}
library(e1071)
```

Build naive Bayes model using the e1071 package
```{r}
nb=naiveBayes(Survived~., data = newtrain, laplace = 1, na.action = na.pass)
```

Apply the model to predicting test data
```{r}
pred=predict(nb, newdata=newtest, type=c("class"))
```

Combine the predictions with the corresponding case ids. 
```{r}
myids=c("PassengerId")
id_col=testset[myids]
newpred=cbind(id_col, pred)
```

Add header to output
```{r}
colnames(newpred)=c("Passengerid", "Survived")
```

Write output to file
```{r}
write.csv(newpred, file="C:/Users/Administrator/Documents/MSDatascience/IST707/Week7/titanic-NB-pred.csv", row.names=FALSE)
```

For more information about naive Bayes in e1071, see the manual at https://cran.r-project.org/web/packages/e1071/e1071.pdf
